TBD

### 2019-09-18 manuscript/manuscript.tex
update repo for Evolutionary Letters submission

\emph{Keywords}: adaptive landscape, ecological networks,
eco-evolutionary dynamics, natural selection, host-parasitoid,
complexity, biodiversity, community context.

@@

\emph{Total Word Count}: 5769

@@

Ecological networks, such as food webs describing who-eats-whom, provide
an explicit representation of the direct and indirect effects that
emerge in a community of interacting species
\citep{Bascompte2014, McCann2012}. Here, we integrate ecological
networks and adaptive landscapes to understand how ecological
communities constrain evolutionary change. The direct and indirect
effects of natural selection can be inferred by quantifying the slope
and curvature of the adaptive landscape \citep{Arnold1992}. For example,
the slope is determined by directional selection gradients acting on
each phenotypic trait and influences the trajectory of evolutionary
change \citep{Lande1979, Arnold1992}. Selective constraints on evolution
increase with the number of traits under directional selection, as this
diminishes the number of optimal solutions \citep{Arnold2003}. The
curvature of the adaptive landscape is governed by nonlinear selection
gradients and acts to indirectly constrain evolution through its effect
on genetic constraints \citep{Arnold1992, Hansen2008}. Genetic
constraints are largely governed by a population's \textbf{G}-matrix
---the additive genetic variances and covariances between traits
\citep{Hansen2008}. Stabilizing selection acts to erode genetic variance
in traits, which can impose a constraint on the ability of those traits
to respond to future selection \citep{Hansen2008}. Correlational
selection may alter the genetic covariance between traits, which may
facilitate or hinder future adaptation depending on pattern of selection
on those traits \citep{Hansen2008}. If we want to predict how ecological
communities shape phenotypic evolution, we must understand how
ecological networks shape the adaptive landscape.

@@

\begin{figure}
\centering
\includegraphics{../analyses/complex_simple_foodwebs_v3.jpeg}
\caption{\label{fig:Conceptual}Experimental manipulation of food-web
structure associated with a leaf-galling midge (\emph{Iteomyia
salicisverruca}) feeding on the willow \emph{Salix hookeriana}. Black
arrows denote the flow of energy in this network of trophic
interactions. In the original food web, we allowed the full suite of egg
and larval parasitoids to impose selection. In the removal food web, we
used mesh bags to exclude the guild of larval parasitoids, only allowing
the egg parasitoid (\emph{Platygaster} sp.) to impose selection. Note
that larval parasitoids also impose indirect effects on gall midge
fitness through intraguild predation on the egg parasitoid. Larval
parasitoids include the following species (from left to right):
\emph{Mesopolobus} sp. (Family: Pteromalidae); \emph{Tetrastichus} sp.
(Family: Eulophidae); and \emph{Torymus} sp. (Family: Torymidae).}
\end{figure}

@@

We setup our food-web manipulation across 128 plants soon after galls
began developing on willows in early June of 2013. These 128 plants came
from eight different plant genotypes that spanned the range of trait
variation observed in this willow population \citep{Barbour2015}. For
the original food web (eight replicates per genotype), we used flagging
tape to mark 14 galled leaves per plant (\textasciitilde{}30 larvae),
allowing the full suite of egg and larval parasitoids to impose
selection. Marking galls with flagging tape ensured that all galls had
similar phenology when we collected galls later in the season. For the
removal food web, we enclosed 14 galled leaves with 10x15cm organza bags
(ULINE, Pleasant Prairie, WI, USA) to exclude three parasitoid species
that attack during larval development. This treatment did not exclude
the egg parasitoid \emph{Platygaster} sp., which attacks prior to gall
initiation \citep[larva initiate gall development in Cecidomyiid
midges:][]{Gagne1989}. It was not possible to reciprocally manipulate
parasitoid attack (i.e.~exclude egg parasitoid, but allow larval
parasitoids), because it was not possible to identify midge oviposition
sites prior to gall formation. In late August, we collected marked and
bagged galls from each plant, placed them into 30 mL vials and kept them
in the lab for 4 months at room temperature. We then opened galls under
a dissecting scope and determined whether larvae survived to pupation
(our measure of fitness) or were parasitized. We did not include other
sources of mortality, such as early larval death, as they could
influence the expression of the gall phenotype and confound estimates of
selection. For the food-web treatment that excluded larval parasitoids,
we further restricted our data by removing any incidental instances of
parasitism by a larval parasitoid. This represented less than 3\% of the
observations in this food-web treatment and allowed us to focus our
inferences of selection on those imposed by the egg parasitoid. Our
final dataset contains survival estimates for 1285 larvae from 613 galls
and 111 plants.

@@

Rather than imposing selection, parasitoids may influence the expression
of herbivore traits which could bias measurements of selection. In our
system, it was plausible that parasitoids may influence chamber diameter
by altering larval feeding behavior or killing larvae before they
complete their development. To estimate this potential bias, we subset
our data to only include galls where there was variation in larval
survival within the same gall (i.e.~1 \textgreater{} mean survival
\textgreater{} 0). If we assume that larvae within each gall should have
similar chamber diameters because they come from the same clutch and
experience the same local environment (an assumption our data supports:
gall ID explains 54\% of the variance in chamber diameter), then the
relationship between chamber diameter and larval survival in this data
subset represents the effect of parasitism on trait expression
(i.e.~bias). We used a GLMM with the same structure as described
previously except that we modeled only a linear relationship between
chamber diameter and larval survival (\(\alpha_{\text{Diam}}\)). We
detected a positive bias in both food-web treatments (original
\(\alpha_{\text{Diam}}\)= 0.36 {[}0.05, 0.67{]}; removal
\(\alpha_{\text{Diam}}\)= 0.42 {[}0.01, 0.82{]}), indicating that
unadjusted relationships would overestimate the strength of selection on
chamber diameter. To account for this bias, we substracted our mean
estimates of bias from our estimates with the full dataset prior to
calculating chamber diameter's selection surface and directional
selection gradient.

@@

\subsection{Consumer loss increases selective
constraints}\label{consumer-loss-increases-selective-constraints}

\indent We found that the removal of larval parasitoids increased the
number of gall midge traits under directional selection (3 of 3)
relative to the original food web (1 of 3)(table \ref{Table:Gradients}).
For example, we observed directional selection for smaller clutch sizes
when we excluded larval parasitoids, but there was no evidence of
selection acting on this trait in the original food web (table
\ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}C). This absence of
selection appeared to be a result of conflicting selection pressures
imposed by each guild of parasitoids. Specifically, when we subset our
data to focus on differences between parasitoid guilds, we found that
larval parasitoids actually impose directional selection for larger
clutch sizes (larval parasitoids \(\beta_{\text{Clutch}}\)= 0.13
{[}0.04, 0.24{]}). This conflicting selection is likely due to trait
differences between guilds, as larger clutches may be easier targets for
egg parasitoids to find, but the more complex gall structure may be more
difficult for larval parasitoids to oviposit through.

@@

We also observed clear evidence of directional selection for midges to
avoid ovipositing on plants with high densities of conspecifics when we
excluded larval parasitoids (table \ref{Table:Gradients}; fig.
\ref{fig:UV_Landscape}B); however, the overall magnitude of selection
did not differ between treatments (table \ref{Table:Gradients}). Still,
there was no clear evidence of directional selection on oviposition in
the original food web (table \ref{Table:Gradients}). Chamber diameter
experienced positive directional selection in both food-web treatments
(fig. \ref{fig:UV_Landscape}A). Although the magnitude of selection on
chamber diameter was relatively higher in the original food web (table
\ref{Table:Gradients}), this was not due to any difference in the
relationship between chamber diameter and survival (selection surfaces:
contrast \(\alpha_{\text{Diam}}\)= 0.04 {[}0.55, -0.5{]}), but was
simply a consequence of the (expected) lower survival of gall midges in
the original food web (contrast \(\bar W\)= 0.27 {[}0.11, 0.42{]}). We
expect this difference to be transient though, since egg parasitoids
would increase in abundance once they are released from intraguild
predation, which would make the strength of selection on gall diameter
comparable to the original food web (if removal \(\bar W\)=original
\(\bar W\), then contrast \(\beta_{\text{Diam}}\)= -0.06 {[}-0.2,
0.1{]}). It is worth noting that positive selection on chamber diameter
in both treatments was unexpected. For example, the fact that larger
galls provide more of a refuge from larval parasitoids makes sense since
they attack after the gall is formed; however, egg parasitoids attack
prior to gall formation, which suggests that chamber diameter is
strongly correlated with an unmeasured trait that is under selection
(e.g.~immune response).

@@

\begin{figure}
\centering
\includegraphics{../analyses/UV_landscapes.pdf}
\caption{\label{fig:UV_Landscape}Adaptive landscape of gall midge
phenotypes in the original food web and with the removal of larval
parasitoids. Each panel corresponds to a different phenotypic trait:
chamber diameter (A); oviposition preference (B); and clutch size (C).
Bold lines represent selection experienced in the original (orange) and
removal (blue) food webs (orange). Thin lines represent bootstrapped
replicates to show the uncertainty in selection. For clarity, we only
display 100 bootstraps even though inferences are based on 1,000
replicates. Note that mean larval survival is plotted on a natural log
scale to reflect the mathematical definition of the adaptive landscape.}
\end{figure}

To visualize the multivariate effects of these selective constraints, we
plotted the adaptive landscape for each trait combination in each
treatment (fig. \ref{fig:MV_Landscape}). The arrows in fig.
\ref{fig:MV_Landscape} represent mean estimates of directional selection
gradients, while contours represent predicted survival of the mean
phenotype in each food-web treatment. Notice that arrows point more
toward a corner of the adaptive landscape for each combination of traits
with the removal of larval parasitoids compared to the original food
web. This indicates that the loss of consumers more strongly favored a
specific combination of traits, rather than allowing for multiple trait
combinations to have comparable fitness.

\begin{figure}
\centering
\includegraphics{../analyses/MV_landscapes.pdf}
\caption{\label{fig:MV_Landscape}Two dimensional view of adaptive
landscapes of gall midge phenotypes in the original food web and with
the removal of larval parasitoids. Each panel corresponds to a different
combination of phenotypic traits: clutch size and chamber diameter (A);
clutch size and oviposition preference (B); oviposition preference and
chamber diameter (C). Arrows represent mean estimates of directional
selection gradients, while contours represent predicted larval survival
of the mean phenotype in each food-web treatment. Notice that arrows
point more toward a corner of the adaptive landscape for each
combination of traits with the removal of larval parasitoids compared to
the original food web. This indicates that the loss of consumers more
strongly favored a specific combination of traits. Note that mean larval
survival is plotted on a natural log scale to reflect the mathematical
definition of the adaptive landscape.}
\end{figure}

\subsection{Consumer loss indirectly increases genetic
constraints}\label{consumer-loss-indirectly-increases-genetic-constraints}

\indent The curvature of the adaptive landscape indirectly affects
genetic constraints and is influenced by directional, quadratic, and
correlational selection gradients. There was no clear evidence of
correlational selection for any combination of traits in either food-web
treatment (table \ref{Table:Gradients}). Similarly, there was no clear
evidence of quadratic selection on chamber diameter or clutch size in
either food-web treatment (table \ref{Table:Gradients}; fig.
\ref{fig:UV_Landscape}A,C). In contrast, our food-web treatment did
alter quadratic selection acting on oviposition preference (table
\ref{Table:Gradients}). The negative relationship between oviposition
preference and larval survival dampened at high densities in the
original food web, but not with the exclusion of larval parasitoids
(fig. \ref{fig:UV_Landscape}B). This dampened relationship was partly
due to a trend for nonlinear selection imposed by larval parasitoids
(\(\gamma_{\text{Pref:Pref}}\)= 0.18 {[}-0.02, 0.42{]}), but was also
magnified by the lower average survival in complex food webs. This was
likely a result of larval parasitoids imposing greater mortality on egg
parasitoids at high gall midge densities (see \textbf{Selection on egg
parasitoids} section), and thus a less than additive effect on gall
midges.

@@

Interestingly, when we translate the effect of the curvature matrix onto
genetic constraints in the next generation, we see that the loss of
consumers had a general tendency to decrease evolvability (fig.
\ref{fig:Evolvability}). Specifically, the removal food web decreased
the average evolvability of 71\% of the random G-matrices in our
simulation. If anything, we expect that this underestimates the true
effect of our removal treatment. For example, if we assume egg
parasitoids would eventually impose similar impact on mean fitness once
they are released from intraguild predation (i.e.~removal
\(\bar W\)=original \(\bar W\)), then the removal food web decreases the
average evolvability in 78\% of the G-matrix scenarios.

\begin{figure}
\centering
\includegraphics{../analyses/delta_evolvability.pdf}
\caption{\label{fig:Evolvability}Change in average evolvability for
10,000 random G-matrices using our best (mean) estimate of the curvature
matrix for each food-web treatment. We found that the curvature of the
removal food web decreased evolvability in 71\% of the G-matrices
(i.e.~the change in evolvability was negative for 71\% of the
simulations), suggesting that the loss of consumers tended to decrease
evolutionary potential of traits in our study.}
\end{figure}

@@

The removal of larval parasitoids not only had direct effects on gall
midge fitness, but also imposed indirect effects that would be felt in
the next generation. For example, the removal of larval parasitoids
altered the relationship between gall midge preference and the
probability of observing egg parasitoids (-0.46 {[} -1.07, -0.02{]} \},
Appendix A), such that the impact of larval parasitoids increased
nonlinearly with higher gall midge densities (fig.
\ref{fig:EggPtoid_Selection}).

@@

We found that the removal of larval parasitoids constrained phenotypic
evolution in gall midges in two key ways. First, more traits contributed
to the slope of the adaptive landscape in the absence of larval
parasitoids, suggesting greater constraints on the trajectory of
phenotypic evolution. Second, excluding larval parasitoids altered the
curvature of the adaptive landscape in such a way that tended to
decrease the evolvability of associated traits. Assuming these traits
have other ecological functions, then this decrease in evolvability
could constrain the gall midge's adaptive potential in the face of novel
selection pressures. Our experiment also revealed evidence of indirect
selection pressures, suggesting that the loss of consumers may have
complex effects on the trajectories of phenotypic evolution. Taken
together, our study provides experimental evidence from the field that
the loss of consumers may constrain the adaptive potential of remaining
populations.

The generality of our results likely depends on the relative abundance
and functional differences between consumers in a community. For
example, if consumers do not differ from each other, then we do not
expect the loss of consumers to modify selective constraints. Also, many
consumers may be at too low of abundances to impose selection on their
resources. Rank abundance curves \citep{Preston1948} and the
disproportionate number of weak interactions in diverse communities
\citep{Paine1992} support this notion. This logic suggests that we may
not have observed the effects we did if we had only removed one larval
parasitoid, because each species had relatively low abundance
\citep{Barbour2016} and they likely share similar ecological roles. When
consumers are functionally different and abundant though, the effect of
consumer loss will depend on whether different species impose
conflicting selection pressures or select for distinct traits. When
consumers impose conflicting selection on traits, as in our study and
others \citep{Weis1985, Abrahamson1997, Start2016, Start2018}, then
consumer diversity acts to neutralize selection and relax selective
constraints. On the other hand, different consumers may impose selection
on different traits; therefore, a more diverse consumer community may
favor a particular combination of traits and increase selective
constraints. Examples of this include strong genetic covariances in
plant resistance to different insect herbivores
\citep{Maddox1990, Wise2007, Wise2013}, although there are also examples
where these covariances are weak \citep{Roche1997, Barbour2015}, or vary
from year-to-year \citep{Johnson2007}. We suggest that gaining
predictive insight to the evolutionary consequences of food-web
disassembly requires an understanding of the mechanisms governing the
assembly of trophic interactions \citep{Bascompte2009}.

@@

An important caveat of our study is that we did not due a factorial
manipulation of both parasitoid guilds, making it difficult to conclude
whether our results would change if we manipulated the presence/absence
of the dominant egg parasitoid. If we assume that higher-order
interactions \citep{Levine2017} are weak between parasitoid guilds, then
we can gain insight to how the loss of the egg parasitoid would alter
selection by isolating the contribution of larval parasitoids to
selection in our original food-web treatment. When we do this, we see
the same qualitative effects as we do when we removed larval
parasitoids. For example, we see clear evidence of all three traits
being under directional selection (i.e.~greater selective constraints,
Appendix A) as well as a decrease, albeit smaller, in trait evolvability
under different G-matrix scenarios (57\%, Appendix A). This suggests
that our results could be robust to this caveat, which was simply not
possible to manipulate given the biology of our system (see
\emph{Manipulating Food-web Structure} section for explanation).

@@

We thank the staff of Humboldt Bay National Wildlife Refuge (U.S. Fish
and Wildlife Service) for facilitating experimental logistics. For
assistance with fieldwork, we thank Ruthie Espanol and Andrew MacDonald.
For funding support, we thank the University of British Columbia (James
Robert Thompson Fellowship and Four-Year Fellowship to M.A.~Barbour),
NSERC (Discovery grant to Greg Crutsinger), and the Swiss National
Science Foundation (grant 31003A\_160671 to J. Bascompte).

@@

\section{Appendix A}\label{appendix-a}

\begin{table}[h]
\caption{Standardized selection gradients acting on egg parasitoids (\textit{Platygaster} sp.)}
\label{Table:ExtendedGradients}
\centering
\begin{tabular}{lc}
\\ 
\hline
\textbf{Selection gradient} & \textbf{Contrast = Original - Removal}  \\ 
\hline
$\beta_{\text{Diam}}$ &

-0.03 [
-0.3,
0.25] \\

$\beta_{\text{Clutch}}$ &

0.07 [
-0.26,
0.39] \\

$\beta_{\text{Pref}}$ &

-0.25 [
-0.64,
0.09] \\

$\gamma_{\text{Diam:Diam}}$ &

-0.05 [
-0.43,
0.33] \\

$\gamma_{\text{Clutch:Clutch}}$ &

-0.21 [
-0.68,
0.26] \\

$\gamma_{\text{Pref:Pref}}$ &

\textbf{
-0.46 [
-1.07,
-0.02] }\\

$\gamma_{\text{Diam:Clutch}}$ &

0 [
-0.29,
0.27] \\

$\gamma_{\text{Diam:Pref}}$ &

0.25 [
-0.04,
0.6] \\

$\gamma_{\text{Clutch:Pref}}$ &

-0.18 [
-0.52,
0.12] \\ 
\hline
\end{tabular}
\bigskip{}
\\
{\footnotesize Note: Values in brackets represent 95\% confidence intervals. Bold values indicate that the 95\% CI does not overlap zero. $\beta_{\text{Diam}}$ has been adjusted for bias.}
\end{table}

\begin{table}[h]
\caption{Standardized selection gradients imposed by larval parasitoids on gall midges in the original food web. This gives insight to selection acting on gall midges in the absence of the dominant egg parasitoid.}
\label{Table:LarvalGradients}
\centering
\begin{tabular}{lc}
\\ 
\hline
\textbf{Selection gradient} & \textbf{Larval Parasitoids} \\ % & \textbf{Original} 
\hline
$\beta_{\text{Diam}}$ & 
%\textbf{
%0.34 [
%0.22,
%0.48] } & 
\textbf{
0.23 [
0.13,
0.36] } \\

$\beta_{\text{Clutch}}$ & 
%0.06 [
%-0.05,
%0.17] & 
\textbf{
0.13 [
0.04,
0.24] }\\

$\beta_{\text{Pref}}$ &
%-0.13 [
%-0.29,
%0.05] & 
\textbf{
-0.17 [
-0.34,
-0.03] } \\

$\gamma_{\text{Diam:Diam}}$ &
%0.13 [
%-0.06,
%0.33] &

0.06 [
-0.07,
0.2]  \\

>>>>

\emph{Keywords}: adaptive landscape, ecological networks,
eco-evolutionary dynamics, natural selection, host-parasitoid, community
context, extinction.

@@

\emph{Total Word Count}: 5833

@@

Ecological networks, such as food webs describing who-eats-whom, provide
an explicit representation of the direct and indirect effects that
emerge in a community of interacting species
\citep{Bascompte2014, McCann2012}. Here, we integrate ecological
networks and adaptive landscapes to understand how ecological
communities constrain evolutionary change. The direct and indirect
effects of natural selection can be inferred by quantifying the slope
and curvature of the adaptive landscape \citep{Arnold1992}. For example,
the slope is determined by directional selection gradients acting on
each phenotypic trait and influences the trajectory of evolutionary
change \citep{Lande1979, Arnold1992}. Selective constraints on evolution
increase with the number of traits under directional selection, as this
diminishes the number of optimal solutions \citep{Arnold2003}. The
curvature of the adaptive landscape is governed by nonlinear selection
gradients and acts to indirectly constrain evolution through its effect
on genetic constraints \citep{Arnold1992, Hansen2008}. Genetic
constraints are largely governed by a population's G-matrix ---the
additive genetic variances and covariances between traits
\citep{Hansen2008}. Stabilizing selection acts to erode genetic variance
in traits, which can impose a constraint on the ability of those traits
to respond to future selection \citep{Hansen2008}. Correlational
selection may alter the genetic covariance between traits, which may
facilitate or hinder future adaptation depending on the pattern of
selection on those traits \citep{Hansen2008}. If we want to predict how
ecological communities shape phenotypic evolution, we must understand
how ecological networks shape the adaptive landscape.

@@

\begin{figure}
\centering
\includegraphics{../analyses/complex_simple_foodwebs_v4.jpeg}
\caption{\label{fig:Conceptual}Experimental manipulation of food-web
structure associated with a leaf-galling midge (\emph{Iteomyia
salicisverruca}) feeding on the willow \emph{Salix hookeriana}. Black
arrows denote the flow of energy in this network of trophic
interactions. In the original food web, we allowed the full suite of egg
and larval parasitoids to impose selection. In the removal food web, we
used mesh bags to exclude the guild of larval parasitoids, only allowing
the egg parasitoid (\emph{Platygaster} sp.) to impose selection. Note
that larval parasitoids also impose indirect effects on gall midge
fitness through intraguild predation on the egg parasitoid. Larval
parasitoids include the following species (from left to right):
\emph{Mesopolobus} sp. (Family: Pteromalidae); \emph{Tetrastichus} sp.
(Family: Eulophidae); and \emph{Torymus} sp. (Family: Torymidae).}
\end{figure}

@@

We setup our food-web manipulation across 128 plants soon after galls
began developing on willows in early June of 2013. These 128 plants came
from eight different plant genotypes that spanned the range of trait
variation observed in this willow population \citep{Barbour2015}. For
the original food web (eight replicates per genotype), we used flagging
tape to mark 14 galled leaves per plant (\textasciitilde{}30 larvae),
allowing the full suite of egg and larval parasitoids to impose
selection. Marking galls with flagging tape ensured that all galls had
similar phenology when we collected galls later in the season. For the
removal food web, we enclosed 14 galled leaves with 10x15cm organza bags
(ULINE, Pleasant Prairie, WI, USA) to exclude three parasitoid species
that attack during larval development. This treatment did not exclude
the egg parasitoid \emph{Platygaster} sp., which attacks prior to gall
initiation \citep[larva initiate gall development in Cecidomyiid
midges:][]{Gagne1989}. It was not possible to reciprocally manipulate
parasitoid attack (i.e.~exclude egg parasitoid, but allow larval
parasitoids), because it was not possible to identify midge oviposition
sites prior to gall formation. In late August, we collected marked and
bagged galls from each plant, placed them into 30 mL vials and kept them
in the lab for 4 months at room temperature. We then opened galls under
a dissecting scope and determined whether larvae survived to pupation
(our measure of fitness) or were parasitized. We did not include other
sources of mortality in our analyses, such as early larval death, as
they could influence the expression of the gall phenotype and confound
estimates of selection. For the food-web treatment that excluded larval
parasitoids, we further restricted our data by removing any incidental
instances of parasitism by a larval parasitoid. This represented less
than 3\% of the observations in this food-web treatment and allowed us
to focus our inferences of selection on those imposed by the egg
parasitoid. Our final dataset contains survival estimates for 1285
larvae from 613 galls and 111 plants.

@@

Rather than imposing selection, parasitoids may influence the expression
of herbivore traits which could bias measurements of selection. In our
system, it was plausible that parasitoids may influence chamber diameter
by altering larval feeding behavior or killing larvae before they
complete their development. To estimate this potential bias, we subset
our data to only include galls where there was variation in larval
survival within the same gall (i.e.~1 \textgreater{} mean survival
\textgreater{} 0). If we assume that larvae within each gall should have
similar chamber diameters because they come from the same clutch and
experience the same local environment (an assumption our data supports:
gall ID explains 54\% of the variance in chamber diameter), then the
relationship between chamber diameter and larval survival in this data
subset represents the effect of parasitism on trait expression
(i.e.~bias). We used a GLMM with the same structure as described
previously except that we modeled only a linear relationship between
chamber diameter and larval survival (\(\alpha_{\text{Diam}}\)). We
detected a positive bias in both food-web treatments (original
\(\alpha_{\text{Diam}}\)= 0.36 {[}0.05, 0.67{]}; removal
\(\alpha_{\text{Diam}}\)= 0.42 {[}0.01, 0.82{]}), indicating that
unadjusted relationships would overestimate the strength of selection on
chamber diameter. To account for this bias, we subtracted our mean
estimates of bias from our estimates with the full dataset prior to
calculating chamber diameter's selection surface and directional
selection gradient.

@@

\subsection{Consumer removal increases selective
constraints}\label{consumer-removal-increases-selective-constraints}

\indent We found that the removal of larval parasitoids increased the
number of gall midge traits under directional selection (3 of 3)
relative to the original food web (1 of 3)(table \ref{Table:Gradients}).
For example, we observed directional selection for smaller clutch sizes
when we removed larval parasitoids, but there was no evidence of
selection acting on this trait in the original food web (fig.
\ref{fig:UV_Landscape}A). This absence of selection appeared to be a
result of conflicting selection pressures imposed by each guild of
parasitoids. Specifically, when we subset our data to focus on
differences between parasitoid guilds, we found that larval parasitoids
actually impose directional selection for larger clutch sizes (larval
parasitoids \(\beta_{\text{Clutch}}\)= 0.13 {[}0.04, 0.24{]}). This
conflicting selection is likely due to trait differences between guilds,
as larger clutches may be easier targets for egg parasitoids to find,
but the more complex gall structure may be more difficult for larval
parasitoids to oviposit through.

@@

We also observed clear evidence of directional selection for midges to
avoid ovipositing on plants with high densities of conspecifics when we
removed larval parasitoids (fig. \ref{fig:UV_Landscape}B); however, the
overall magnitude of selection did not differ between treatments (table
\ref{Table:Gradients}). Still, there was no clear evidence of
directional selection on oviposition in the original food web (table
\ref{Table:Gradients}). Chamber diameter experienced positive
directional selection in both food-web treatments (fig.
\ref{fig:UV_Landscape}C). Although the magnitude of selection on chamber
diameter was relatively higher in the original food web (table
\ref{Table:Gradients}), this was not due to any difference in the
relationship between chamber diameter and survival (selection surfaces:
contrast \(\alpha_{\text{Diam}}\)= 0.04 {[}0.55, -0.5{]}), but was
simply a consequence of the (expected) lower survival of gall midges in
the original food web (contrast \(\bar W\)= 0.27 {[}0.11, 0.42{]}). We
expect this difference to be transient though, since egg parasitoids
would increase in abundance once they are released from intraguild
predation, which would make the strength of selection on gall diameter
comparable to the original food web (if removal \(\bar W\)=original
\(\bar W\), then contrast \(\beta_{\text{Diam}}\)= -0.06 {[}-0.2,
0.1{]}). It is worth noting that positive selection on chamber diameter
in both treatments was unexpected. For example, the fact that larger
galls provide more of a refuge from larval parasitoids makes sense since
they attack after the gall is formed; however, egg parasitoids attack
prior to gall formation, which suggests that chamber diameter is
strongly correlated with an unmeasured trait that is under selection
(e.g.~immune response).

@@

\begin{figure}
\centering
\includegraphics{../analyses/UV_landscapes.pdf}
\caption{\label{fig:UV_Landscape}Adaptive landscape of gall midge
phenotypes in the original food web and with the removal of larval
parasitoids. Each panel corresponds to a different phenotypic trait:
clutch size (A); oviposition preference (B); and chamber diameter (C).
Bold lines represent selection experienced in the original (orange) and
removal (blue) food webs. Thin lines represent bootstrapped replicates
to show the uncertainty in selection. For clarity, we only display 100
bootstraps even though inferences are based on 1,000 replicates. Note
that mean larval survival is plotted on a natural log scale to reflect
the mathematical definition of the adaptive landscape.}
\end{figure}

To visualize the multivariate effects of these selective constraints, we
plotted the adaptive landscape for each trait combination in each
treatment (fig. \ref{fig:MV_Landscape}). The arrows in fig.
\ref{fig:MV_Landscape} represent mean estimates of directional selection
gradients, while contours represent predicted survival of the mean
phenotype in each food-web treatment. Notice that arrows point more
toward a corner of the adaptive landscape for each combination of traits
with the removal of larval parasitoids compared to the original food
web. This indicates that the removal of consumers more strongly favored
a specific combination of traits, rather than allowing for multiple
trait combinations to have comparable fitness.

\begin{figure}
\centering
\includegraphics{../analyses/MV_landscapes.pdf}
\caption{\label{fig:MV_Landscape}Two dimensional view of adaptive
landscapes of gall midge phenotypes in the original food web and with
the removal of larval parasitoids. Each panel corresponds to a different
combination of phenotypic traits: clutch size and chamber diameter (A);
clutch size and oviposition preference (B); oviposition preference and
chamber diameter (C). Arrows represent mean estimates of directional
selection gradients, while contours represent predicted larval survival
of the mean phenotype in each food-web treatment. Notice that arrows
point more toward a corner of the adaptive landscape for each
combination of traits with the removal of larval parasitoids compared to
the original food web. This indicates that the removal of consumers more
strongly favored a specific combination of traits. Note that mean larval
survival is plotted on a natural log scale to reflect the mathematical
definition of the adaptive landscape.}
\end{figure}

\subsection{Consumer removal indirectly increases genetic
constraints}\label{consumer-removal-indirectly-increases-genetic-constraints}

\indent The curvature of the adaptive landscape indirectly affects
genetic constraints and is influenced by directional, quadratic, and
correlational selection gradients. There was no clear evidence of
correlational selection for any combination of traits in either food-web
treatment (table \ref{Table:Gradients}). Similarly, there was no clear
evidence of quadratic selection on chamber diameter or clutch size in
either food-web treatment (table \ref{Table:Gradients}). In contrast,
our food-web treatment did alter quadratic selection acting on
oviposition preference (table \ref{Table:Gradients}). The negative
relationship between oviposition preference and larval survival dampened
at high densities in the original food web, but not with the removal of
larval parasitoids (fig. \ref{fig:UV_Landscape}B). This dampened
relationship was partly due to a trend for nonlinear selection imposed
by larval parasitoids (\(\gamma_{\text{Pref:Pref}}\)= 0.18 {[}-0.02,
0.42{]}), but was also magnified by the lower average survival in the
original food web. This was likely a result of larval parasitoids
imposing greater mortality on egg parasitoids at high gall midge
densities (see \textbf{Selection on egg parasitoids} section), and thus
a less than additive effect on gall midges.

@@

Interestingly, when we translate the effect of the curvature matrix onto
genetic constraints in the next generation, we see that the removal of
consumers had a general tendency to decrease evolvability (fig.
\ref{fig:Evolvability}). Specifically, the removal food web decreased
the average evolvability of 71\% of the random G-matrices in our
simulation. If anything, we expect that this underestimates the true
effect of our removal treatment. For example, if we assume egg
parasitoids would eventually impose similar impact on mean fitness once
they are released from intraguild predation (i.e.~removal
\(\bar W\)=original \(\bar W\)), then the removal food web decreases the
average evolvability in 78\% of the G-matrix scenarios.

\begin{figure}
\centering
\includegraphics{../analyses/delta_evolvability.pdf}
\caption{\label{fig:Evolvability}Change in average evolvability for
10,000 random G-matrices using our best (mean) estimate of the curvature
matrix for each food-web treatment. We found that the curvature of the
removal food web decreased evolvability in 71\% of the G-matrices
(i.e.~the change in evolvability was negative for 71\% of the
simulations), suggesting that the removal of consumers tended to
decrease evolutionary potential of traits in our study.}
\end{figure}

@@

The removal of larval parasitoids not only had direct effects on gall
midge fitness, but also imposed indirect effects that would be felt in
the next generation. For example, the removal of larval parasitoids
altered the relationship between gall midge preference and the
probability of observing egg parasitoids
(\(\gamma_{\text{Pref:Pref}}\)=-0.46 {[} -1.07, -0.02{]}, table S1),
such that the impact of larval parasitoids increased nonlinearly with
higher gall midge densities (fig. \ref{fig:EggPtoid_Selection}).

@@

We found that the removal of larval parasitoids constrained phenotypic
evolution in gall midges in two key ways. First, more traits contributed
to the slope of the adaptive landscape in the absence of larval
parasitoids, suggesting greater constraints on the trajectory of
phenotypic evolution. Second, removing larval parasitoids altered the
curvature of the adaptive landscape in such a way that tended to
decrease the evolvability of associated traits. Assuming these traits
have other ecological functions, then this decrease in evolvability
could constrain the gall midge's adaptive potential in the face of novel
selection pressures. Our experiment also revealed evidence of indirect
selection pressures, suggesting that the loss of consumers may have
complex effects on the trajectories of phenotypic evolution. Taken
together, our study provides experimental evidence from the field that
the loss of consumers may constrain the adaptive potential of remaining
populations.

The generality of our results likely depends on the relative abundance
and functional differences between consumers in a community. For
example, if consumers do not differ from each other, then we do not
expect the loss of consumers to modify selective constraints. Also, many
consumers may be at too low of abundances to impose selection on their
resources. Rank abundance curves \citep{Preston1948} and the
disproportionate number of weak interactions in diverse communities
\citep{Paine1992} support this notion. This logic suggests that we may
not have observed the effects we did if we had only removed one larval
parasitoid, because each species had relatively low abundance
\citep{Barbour2016} and they likely share similar ecological roles. When
consumers are functionally different and abundant though, the effect of
consumer loss will depend on whether different consumers impose
conflicting selection pressures or select for distinct traits. When
consumers impose conflicting selection on traits, as in our study and
others \citep{Weis1985, Abrahamson1997, Start2016, Start2018}, then
consumer diversity acts to neutralize selection and relax selective
constraints. On the other hand, different consumers may impose selection
on different traits; therefore, a more diverse consumer community may
favor a particular combination of traits and increase selective
constraints. Examples of this include strong genetic covariances in
plant resistance to different insect herbivores
\citep{Maddox1990, Wise2007, Wise2013}, although there are also examples
where these covariances are weak \citep{Roche1997, Barbour2015}, or vary
from year-to-year \citep{Johnson2007}. We suggest that gaining
predictive insight to the evolutionary consequences of food-web
disassembly requires an understanding of the mechanisms governing the
assembly of trophic interactions \citep{Bascompte2009}.

@@

An important caveat of our study is that we did not do a factorial
manipulation of both parasitoid guilds, making it difficult to conclude
whether our results would change if we manipulated the presence/absence
of the dominant egg parasitoid. If we assume that higher-order
interactions \citep{Levine2017} are weak between parasitoid guilds, then
we can gain insight to how the loss of the egg parasitoid would alter
selection by isolating the contribution of larval parasitoids to
selection in our original food-web treatment. When we do this, we see
the same qualitative effects as we do when we removed larval
parasitoids. For example, we see clear evidence of all three traits
being under directional selection (i.e.~greater selective constraints,
table S2) as well as a decrease, albeit smaller, in trait evolvability
under different G-matrix scenarios (57\%, fig. S1). This suggests that
our results could be robust to this caveat, which was simply not
possible to manipulate given the biology of our system (see
\textbf{Manipulating Food-web Structure} section for explanation).

@@

We thank the staff of Humboldt Bay National Wildlife Refuge (U.S. Fish
and Wildlife Service) for facilitating experimental logistics. For
assistance with fieldwork, we thank Ruthie Espanol and Andrew MacDonald.
For funding support, we thank the University of British Columbia (James
Robert Thompson Fellowship and Four-Year Fellowship to M.A.~Barbour),
NSERC (Discovery grant to Greg Crutsinger), and the Swiss National
Science Foundation (grant 31003A\_160671 to J. Bascompte). The authors
declare no conflict of interest.

@@
